---
title: "DivEMT: Neural Machine Translation Post-Editing Effort Across Typologically Diverse Languages"
author: "Antonio Toral"
date: "Generation date: `r format(Sys.time(), '%b %d, %Y - %H:%M:%S')`"
output: 
  html_document:
    toc: true
    code_folding: show
    toc_float: 
        collapsed: false
        smooth_scroll: true
    number_sections: true
---


This document contains the step-by-step statistical analyses of the post-editing (PE) effort experiments described in:
TBD (2022). DivEMT: Assessing Neural Machine Translation Post-Editing Effort
across Typologically Diverse Languages.

This is partially based on a previous analysis for the publication:
Toral, A., Wieling, M., and Way, A. (2018, in press). Post-editing effort of a novel with statistical and neural machine translation. Frontiers in Digital Humanities.

Variables used

- Predictors
  - Fixed
    - Length of the source sentence (characters better than words according to the 2018 paper)
    - task_type: the translation condition. It is a factor with 3 levels: HT (human translation from scratch), PE1 (post-editing the output of Google Translate) and PE2 (post-editing the output of mBART50). The reference level is HT.
    - trial number
    - language (6 languages)
  - Random factors
    - Subject (18 translators)
    - Item (nested). Document (107 documents) / Sentence (413 sentences)

- Dependent variables
  - Translation time in seconds (temporal effort)
  - number of keystrokes (technical effort)
  - number of pauses | avg length of pauses | pause to total time ratio (cognitive effort)
  
- Hypothesised interactions
  - trial * task_type: does the longitudinal effect depend on translation type?
  - lang_id * task_type: does the effect of translation type vary per language?
  - len_sl_char * task_type: does translation speed of sentences of different length depend on translation type? Some translators said post-editing is useful only for short sentences. In the 2018 paper post-editing of NMT (RNN) was less effective for long sentences.

- Hypothesised random slopes
  - 1+trial|subject -> longitudinal effect depends on the subject
  - 1+task_type|subject -> effect of translation mode depends on subject. MT suggestions may help some translators more than others.
  - 1+task_type|item -> effect of translation mode depends on sentence. MT quality varies across sentences.

# Load libraries and install required packages if not installed yet
```{r message=FALSE}
# install packages if not there yet
packages <- c("car", "effects", "ggplot2", "jtools", "lme4", "lmerTest", "mgcv", "optimx", "tidyr")
if (length(setdiff(packages, rownames(installed.packages()))) > 0) {
    install.packages(setdiff(packages, rownames(installed.packages())), repos = "http://cran.us.r-project.org")
}

library(car)
library(effects)
library(ggplot2)
library(jtools)
library(lme4)
library(mgcv)
library(optimx)
library(interactions)
```

```{r versioninfo}
# display version information
R.version.string
packageVersion('mgcv')
packageVersion('car')
packageVersion('lme4')
packageVersion('mgcv')
packageVersion('optimx')
```

# Load dataset
```{r}
setwd("~/projects/divemt/data/processed")
# Save the filtered_data variable produced in the analysis.ipynb notebook to load it here.
mydf <- read.csv("filtered_divemt.tsv", sep="\t")
# 7434 rows (18 translators * 413 sentences)
# Outliers have been already removed from this dataset (17 sentences for all languages)
# create additional variables (trial number, time and number of keystrokes per character, seconds per keystrokes and words per hour)
mydf$trial <- mydf$per_subject_visit_order

mydf$time <- mydf$event_time / 1000 # time from ms to seconds
mydf$time_h = mydf$time / 3600 # time from seconds to hours

mydf$time_div_sl_len_char = mydf$time / mydf$src_len_chr # time per character
mydf$k_total_div_sl_len_char = mydf$k_total / mydf$src_len_chr # keystrokes per character
mydf$k_total_div_time = mydf$k_total / mydf$time # keystrokes per second
mydf$sl_len_word_div_time_hours = mydf$src_len_wrd / (mydf$time_h) # words per hour

# scale the continuous predictors
mydf$src_len_chr.s <- scale(mydf$src_len_chr)
mydf$src_len_chr.s = c(mydf$src_len_chr.s)
mydf$trial.s <- scale(mydf$trial)
mydf$trial.s = c(mydf$trial.s)

# relevel predictor language. Use as reference the slowest one (Ukrainian)
mydf$lang_id <- as.factor(mydf$lang_id)
mydf$lang_id <- relevel(mydf$lang_id, "ukr")

# document_id as a factor, since it'll be used as a random factor
mydf$document_id <- as.factor(mydf$doc_id)
```

# Overall Relative Changes (PE with MT1 and MT2 compared to HT)

Productivity (words processed per hour)

```{r}
product_per_task_type <- tapply(mydf$src_len_wrd, mydf$task_type, FUN=sum) / tapply(mydf$time/3600, mydf$task_type, FUN=sum) #productivity: words/hour
product_per_task_type
for (i in seq(2,3)){
  print((product_per_task_type[i] - product_per_task_type[1]) / product_per_task_type[1])
}
```

Relative changes in productivity: MT1 62%, MT2 30% (in IT: MT1 101%, MT2 74%)

Temporal effort (seconds per source character)

```{r}
temp_eff_per_task_type <- tapply(mydf$time, mydf$task_type, FUN=sum) / tapply(mydf$src_len_chr, mydf$task_type, FUN=sum) # temp effort: seconds/character
temp_eff_per_task_type
for (i in seq(2,3)){
  print((temp_eff_per_task_type[i] - temp_eff_per_task_type[1]) / temp_eff_per_task_type[1])
}
```

Temporal reduction: MT1 -38%, MT2 -23% (in IT: MT1 -51%, MT2 -43%)

Technical effort (keystrokes per source character)

```{r}
tech_eff_per_task_type <- tapply(mydf$k_total, mydf$task_type, FUN=sum) / tapply(mydf$src_len_chr, mydf$task_type, FUN=sum) # technical effort: keystrokes/character
tech_eff_per_task_type
for (i in seq(2,3)){
  print((tech_eff_per_task_type[i] - tech_eff_per_task_type[1]) / tech_eff_per_task_type[1])
}
```

Keystroke reduction: MT1 -55%, MT2 -39% (in IT MT1 -77%, MT2 -70%)

Typing speed (keystrokes per second)

```{r}
type_speed_per_task_type <- tapply(mydf$k_total, mydf$task_type, FUN=sum) / tapply(mydf$time, mydf$task_type, FUN=sum) # keystrokes/second
type_speed_per_task_type
for (i in seq(2,3)){
  print((type_speed_per_task_type[i] - type_speed_per_task_type[1]) / type_speed_per_task_type[1])
}
```

MT1 -27%, MT2 -21% (in IT MT1 -55%, MT2 -48%)

# Temporal Effort

## Outliers

```{r}
# 1 boxplot per subject and task
par(mfrow=c(1,2), family  = "Arial")
boxplot(mydf$time ~ mydf$subject_id + mydf$task_type)
boxplot(mydf$time_div_sl_len_char ~ mydf$subject_id + mydf$task_type)#, ylim=c(0, 40))

# 1 boxplot per task and subject
par(mfrow=c(1,2), family  = "Arial")
boxplot(mydf$time ~ mydf$task_type + mydf$subject_id)
boxplot(mydf$time_div_sl_len_char ~ mydf$task_type + mydf$subject_id)#, ylim=c(0, 40))

# 1 boxplot per subject
par(mfrow=c(1,2), family  = "Arial")
boxplot(mydf$time ~ mydf$subject_id)
boxplot(mydf$time_div_sl_len_char ~ mydf$subject_id)#, ylim=c(0, 40))

# 1 boxplot per task
par(mfrow=c(1,2), family  = "Arial")
boxplot(mydf$time ~ mydf$task_type)
boxplot(mydf$time_div_sl_len_char ~ mydf$task_type)#, ylim=c(0, 40))
```

No clear outliers, e.g. sentences for which translators take more than 6 seconds per source character. Outliers have been already removed, based on edit time > 45 min

## Dependent Variable Transformation

The dependent variable (translation time) has a very long right tail. Hence we transform it logarithmically.

```{r}
mydf$time.l <- log(mydf$time)
par(mfrow=c(1,2), family  = "Arial")
plot(density(mydf$time))
plot(density(mydf$time.l))
```

## LMER: Nested Random Effect

```{r}
te.lmer0 = lmer(time.l ~ src_len_chr.s + (1|subject_id) + (1|item_id), mydf, REML=T)
summary(te.lmer0)

te.lmer0b = lmer(time.l ~ src_len_chr.s + (1|subject_id) + (1|document_id/item_id), mydf, REML=T)
summary(te.lmer0b)
AIC(te.lmer0) - AIC(te.lmer0b) # 46.5 (nested random effect makes sense)
```

## LMER: Fixed Effects

```{r}
te.lmer0b = lmer(time.l ~ src_len_chr.s + (1|subject_id) + (1|document_id/item_id), mydf, REML=F)

te.lmer1 = lmer(time.l ~ src_len_chr.s + task_type + (1|subject_id) + (1|document_id/item_id), mydf, REML=F)
AIC(te.lmer0) - AIC(te.lmer1) # 1123 (>2 -> better model)
summary(te.lmer1) # pe1 & pe2 significantly faster than ht (|t|>2 for task_type)

te.lmer2 = lmer(time.l ~ src_len_chr.s + task_type + lang_id + (1|subject_id) + (1|document_id/item_id), mydf, REML=F)
AIC(te.lmer1) - AIC(te.lmer2) # 4 (>2 -> better model)
summary(te.lmer2) # ara, nld & tur significantly faster than ara. ita and vie are not.

te.lmer3 = lmer(time.l ~ src_len_chr.s + task_type + lang_id + trial.s + (1|subject_id) + (1|document_id/item_id), mydf, REML=F)
AIC(te.lmer2) - AIC(te.lmer3) # -1.44 (<2 -> not better model)
summary(te.lmer3) # not enough evidence for a longitudinal effect: translators do not get significantly faster (|t|<2 for trial)
```

## LMER: Interactions of Fixed Effects

```{r}
te.lmer4a = lmer(time.l ~ src_len_chr.s + lang_id + task_type * trial.s + (1|subject_id) + (1|document_id/item_id), mydf, REML=F)
summary(te.lmer4a)
AIC(te.lmer3) - AIC(te.lmer4a) #14. Translators get faster in PE, but slower in HT!

iplot_te_task_trial <- interact_plot(te.lmer4a, pred = trial.s, modx = task_type)
iplot_te_task_trial + theme(axis.title = element_text(family = "Arial"),
           legend.text = element_text(family = "Arial"),
           legend.title = element_text(family = "Arial"),
           strip.text = element_text(family = "Arial"),
           axis.text.x = element_text(family = "Arial"),
           axis.text.y = element_text(family = "Arial"))
```

The model with the interaction between task_type and trial number is better. The interaction is significant.

Figure: Translators get faster in PE, but slower in HT

```{r}
te.lmer4b = lmer(time.l ~ src_len_chr.s + lang_id * task_type + trial.s + (1|subject_id) + (1|document_id/item_id), mydf, REML=F)
summary(te.lmer4b)
AIC(te.lmer3) - AIC(te.lmer4b) # 164. PE is not equally useful in all languages

iplot_te_task_trial <- cat_plot(te.lmer4b, pred = lang_id, modx = task_type)
iplot_te_task_trial + theme(axis.title = element_text(family = "Arial"),
           legend.text = element_text(family = "Arial"),
           legend.title = element_text(family = "Arial"),
           strip.text = element_text(family = "Arial"),
           axis.text.x = element_text(family = "Arial"),
           axis.text.y = element_text(family = "Arial"))
```

The interaction between tasktype and lang_id leads to a better model.

```{r}
te.lmer4c = lmer(time.l ~ task_type * `src_len_chr.s` + trial.s + (1|subject_id) + (1|document_id/item_id), mydf, REML=F)
summary(te.lmer4c)
AIC(te.lmer3) - AIC(te.lmer4c) #-7

iplot_te_len_task <- interact_plot(te.lmer4c, pred = src_len_chr.s, modx = task_type, y.label = "time (log)", x.label = "character source length (scaled)", legend.main = "condition")
iplot_te_len_task + theme(axis.title = element_text(family = "Arial"),
           legend.text = element_text(family = "Arial"),
           legend.title = element_text(family = "Arial"),
           strip.text = element_text(family = "Arial"),
           axis.text.x = element_text(family = "Arial"),
           axis.text.y = element_text(family = "Arial"))
```

The model with the interaction between task_type and length is worse. The interaction is not significant.

Figure: no difference for different lengths of sentences.

```{r}
te.lmer4d = lmer(time.l ~ src_len_chr.s + lang_id * task_type + task_type * trial.s + (1|subject_id) + (1|document_id/item_id), mydf, REML=F)
summary(te.lmer4d)
AIC(te.lmer4b) - AIC(te.lmer4d) # 10.7
AIC(te.lmer4a) - AIC(te.lmer4d) # 161

iplot_te_task_trial <- cat_plot(te.lmer4b, pred = lang_id, modx = task_type)
iplot_te_task_trial + theme(axis.title = element_text(family = "Arial"),
           legend.text = element_text(family = "Arial"),
           legend.title = element_text(family = "Arial"),
           strip.text = element_text(family = "Arial"),
           axis.text.x = element_text(family = "Arial"),
           axis.text.y = element_text(family = "Arial"))
```

Both significant interactions lead to a better model than either on its own.

## LMER: Random Effects

```{r}
ranef(te.lmer4d)$subject_id
```

We can observe that the adjustments for specific subjects to the intercept are slightly different, positive for e.g. ara_T1 and ita_T1 and negative for e.g. ita_T5.

Now we check whether the addition of random slopes for subject results in a better model.

```{r}
te.lmer4dREML = lmer(time.l ~ src_len_chr.s + lang_id * task_type + task_type * trial.s + (1|subject_id) + (1|document_id/item_id), mydf, REML=T) # We build the previous best model with REML=T so that it can be compared to models with random slopes

#subject
te.lmer5a = lmer(time.l ~ src_len_chr.s + lang_id * task_type + task_type * trial.s + (1|subject_id) + (0+trial.s|subject_id) + (1|document_id/item_id), mydf, REML=T,
                 control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))
AIC(te.lmer4dREML) - AIC(te.lmer5a) #182

te.lmer5b = lmer(time.l ~ src_len_chr.s + lang_id * task_type + task_type * trial.s + (1|subject_id) + (1+trial.s|subject_id) + (1|document_id/item_id), mydf, REML=T,
                 control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))  # TODO does not converge

te.lmer5c = lmer(time.l ~ src_len_chr.s + lang_id * task_type + task_type * trial.s + (1|subject_id) + (1+task_type|subject_id) + (1|document_id/item_id), mydf, REML=T,
                 control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb'))) # TODO does not converge
```

Slope 0+trial.s for subject improves the model.


Now we check whether the addition of random slopes for item results in a better model.

```{r}
te.lmer6a = lmer(time.l ~ src_len_chr.s + lang_id * task_type + task_type * trial.s + (1|subject_id) + (0+trial.s|subject_id) + (1|document_id/item_id) + (0+trial.s|document_id/item_id), mydf, REML=T,
                 control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))
  
AIC(te.lmer5a) - AIC(te.lmer6a) #10


te.lmer6b = lmer(time.l ~ src_len_chr.s + lang_id * task_type + task_type * trial.s + (1|subject_id) + (0+trial.s|subject_id) + (1|document_id/item_id) + (0+task_type|document_id/item_id), mydf, REML=T,
                 control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb'))) # takes very long
AIC(te.lmer5a) - AIC(te.lmer6b) #91

te.lmer6c = lmer(time.l ~ src_len_chr.s + lang_id * task_type + task_type * trial.s + (1|subject_id) + (0+trial.s|subject_id) + (1+task_type|document_id/item_id), mydf, REML=T,
                 control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb'))) # takes very long
AIC(te.lmer6b) - AIC(te.lmer6c) #-78

summary(te.lmer6b) # the best model thus far
```

## LMER Assumptions

```{r}
# heterodasticity
par(mfrow=c(1,1), family  = "Arial")
plot(fitted(te.lmer6b), resid(te.lmer6b)) # OK

# normality
par(mfrow=c(1,1), family  = "Arial")
qqp(resid(te.lmer6b)) # not OK
```

Heterodasticity looks fine but the distribution of residuals deviates from normality.

## Model criticism

```{r}
# Remove outliers
mydfno = mydf[abs(scale(resid(te.lmer6b))) < 2.5,]
dim(mydfno) - dim(mydf)
1 - (dim(mydfno)/dim(mydf))# 2.4% of the data points are removed

# Fit model on data without outliers
library(lmerTest)

te.lmer6bno = lmer(time.l ~ src_len_chr.s + lang_id * task_type + task_type * trial.s + (1|subject_id) + (0+trial.s|subject_id) + (1|document_id/item_id) + (0+task_type|document_id/item_id), mydfno, REML=T,
                 control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))

# Check normality
par(mfrow=c(1,2), family="Arial")
qqp(resid(te.lmer6b))
qqp(resid(te.lmer6bno)) # Normality looks fine in the model without outliers.

summary(te.lmer6b)
summary(te.lmer6bno)
```

Most of the predictors and interactions that were significant remain so in the model without outliers.

trial is not significant (neither on its own nor in an interaction). Therefore we remove it and check if the simpler resulting model is not significantly worse in terms of AIC.

```{r}
te.lmer7bno = lmer(time.l ~ src_len_chr.s + lang_id * task_type + (1|subject_id) + (1|document_id/item_id) + (0+task_type|document_id/item_id), mydfno, REML=T,
                 control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))

AIC(te.lmer6bno) - AIC(te.lmer7bno) # -134
summary(te.lmer7bno)
```

```{r}
ranef(te.lmer7bno)$subject_id
```

## Variance Explained and Significance

Percentage of variance explained by the models with and without outliers
```{r}
cor(mydf$time.l, fitted(te.lmer6b))^2 # 0.56
cor(mydfno$time.l, fitted(te.lmer6bno))^2 # 0.63
cor(mydfno$time.l, fitted(te.lmer7bno))^2 # 0.62
```

Use PE1 as reference level of translation condition, to check if PE2 is significantly different than PE1. So far HT was the reference so we only compared HT-PE1 and HT-PE2

```{r}
mydfno$task_type <- as.factor(mydfno$task_type)
mydfno$task_type <- relevel(mydfno$task_type, "pe1")
te.lmer6bno_pe1 = lmer(time.l ~ src_len_chr.s + lang_id * task_type + task_type * trial.s + (1|subject_id) + (0+trial.s|subject_id) + (1|document_id/item_id) + (0+task_type|document_id/item_id), mydfno, REML=T,
                 control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))

mydfno$task_type <- relevel(mydfno$task_type, "ht")
```

To conclude with temporal effort, we report the time in each condition according to the model without outliers and their relative differences.

```{r}
# UKR (reference level)
exp(4.92022)          #HT:  137 seconds
exp(4.92022 - 0.49480)  #PE1: 83.5
exp(4.92022 - 0.22159)  #PE2: 109.8

# ITA
exp(4.92022-0.552)   #HT:  78.9 seconds
exp(4.92022 - 0.49480 - 0.39861)  #PE1: 56.1
exp(4.92022 - 0.22159 - 0.38659)  #PE2: 74.6
```

## Noise from Subjects vs Languages

Compare the magnitude of noise introduced by different subjects and the differences between languages

With random effects, the model adding language as a nested random intercept is not better.

```{r}
te.lmer_subj = lmer(time.l ~ src_len_chr.s + task_type + (1|subject_id) + (1|document_id/item_id), mydfno, REML=T)
summary(te.lmer_subj)

te.lmer_subj_lang = lmer(time.l ~ src_len_chr.s + task_type + (1|lang_id/subject_id) + (1|document_id/item_id), mydfno, REML=T)

AIC(te.lmer_subj) - AIC(te.lmer_subj_lang) # equivalent -> language as nested random effect not relevant!

cor(mydfno$time.l, fitted(te.lmer_subj))^2
cor(mydfno$time.l, fitted(te.lmer_subj_lang))^2 # same amount of variance explained
```

With fixed effects, subject_id is a more important predictor than language_id.

```{r}
te.lmer_basef = lmer(time.l ~ src_len_chr.s + task_type + (1|document_id/item_id), mydfno, REML=T)
te.lmer_subjf = lmer(time.l ~ src_len_chr.s + task_type + subject_id + (1|document_id/item_id), mydfno, REML=T)
summary(te.lmer_subjf)
te.lmer_langf = lmer(time.l ~ src_len_chr.s + task_type + lang_id + (1|document_id/item_id), mydfno, REML=T)
summary(te.lmer_langf)

AIC(te.lmer_basef) - AIC(te.lmer_subjf) #2737
AIC(te.lmer_basef) - AIC(te.lmer_langf) #1296

cor(mydfno$time.l, fitted(te.lmer_subjf))^2 #0.59
cor(mydfno$time.l, fitted(te.lmer_langf))^2 #0.49
```